Fluid-mechanical erosion creates anisotropic porous media

### 2022-05-07 docs/PhysicsPaper/StokesErosion.tex
NM

% From Nick's Latex Notes
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand {\bq} {\bvec{q}}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\qavg}{\bar{q}}
\newcommand{\pavg}{\bar{p}}

@@

%-----------------------------------------------------------------------------------------------%
\section{Measuring permeability and other porous-media properties}
\label{sec:}
%-----------------------------------------------------------------------------------------------%
Darcy's law is given by
\begin{equation}
\bq = - \bvec{K} \grad \phi
\end{equation}
where $\bq$ is the specific discharge, $\phi = z + p/\gamma$ is the hydraulic head, and $\gamma = \rho g$. The rank-2 tensor $\bvec{K}$ is the {\em conductivity}, which is related to the {\em permeability} $\bvec{k}$ via $\bvec{K} = \bvec{k} \gamma/\mu$, where $\mu$ is the viscosity of the fluid. The units of conductivity are length-per-time (i.e.~velocity) and the units of permeability are length squared.
Perhaps the most standard parameter to characterize the porous medium is the permeability. We eliminate gravity from consideration to get
\begin{equation}
\bq = - \bvec{k} \grad p / \mu
\end{equation}
Henceforth, we set $\mu = 1$ by the non-dimensionalization.

@@

Next, consider the pressure averaged over a vertical slice
\begin{equation}
\pavg(x_0) := \frac{1}{2} \int_{-1}^{1} p(x_0, y) dy
\end{equation}
Integrating \eqref{q1} over the domain $[-x_0, x_0] \times [-1,1]$, applying the fundamental theorem of calculus, and rearranging gives
\begin{equation}
r = \left( \frac{\pavg(-x_0) - \pavg(x_0)}{2 x_0 \qavg} \right)
\end{equation}
This formula gives the resistivity based on measurements of the pressure and discharge. In particular, to compute the resistivity of the porous matrix, we must set $x_0 =1$ in principle. (If a larger value of $x_0$ is chosen, then one would be in effect averaging the porous-medium resistivity with the vanishing resistivity of the empty buffer region.)

>>>>

% From Nick's Latex Notes
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand {\bq} {\bvec{q}}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\qavg}{\bar{q}}
\newcommand{\pavg}{\bar{p}}
\newcommand{\pup}{p_u}
\newcommand{\pdn}{p_d}
\newcommand {\bu} {\bvec{u}}
\newcommand{\nhat}{\hat{\bvec{n}}}
\newcommand{\shat}{\hat{\bvec{s}}}
\newcommand{\FD}{\bvec{F_D}}
\newcommand{\ex}{\hat{\bvec{e}}_1}

@@

%-----------------------------------------------------------------------------------------------%
\section{Measuring permeability and other porous-media properties}
\label{sec:}
%-----------------------------------------------------------------------------------------------%

%-----------------------------------------------------------------------------------------------%
\subsection{Resistivity and permeability}
Darcy's law is given by
\begin{equation}
\bq = - \bvec{K} \grad \phi
\end{equation}
where $\bq$ is the specific discharge, $\phi = z + p/\gamma$ is the hydraulic head, and $\gamma = \rho g$. The rank-2 tensor $\bvec{K}$ is the {\em conductivity}, which is related to the {\em permeability} $\bvec{k}$ via $\bvec{K} = \bvec{k} \gamma/\mu$, where $\mu$ is the viscosity of the fluid. The units of conductivity are length-per-time (i.e.~velocity) and the units of permeability are length squared.
Perhaps the most standard parameter to characterize the porous medium is the permeability. We eliminate gravity from consideration to get
\begin{equation}
\bq = - \bvec{k} \grad p / \mu
\end{equation}
Henceforth, we set $\mu = 1$ by the non-dimensionalization.

@@

Next, consider the pressure averaged over a vertical slice
\begin{equation}
\pavg(x_0) := \frac{1}{2} \int_{-1}^{1} p(x_0, y) dy
\end{equation}
In particular, the upstream and downstream pressures are respectively
\begin{equation}
\pup = \pavg(-1) \, , \quad \pdn = \pavg(1)
\end{equation}

Integrating \eqref{q1} over the porous-medium domain $[-1, 1] \times [-1,1]$, applying the fundamental theorem of calculus, and rearranging gives
\begin{equation}
r = \left( \frac{\pup - \pdn}{2 \qavg} \right)
\end{equation}
This exact formula gives the resistivity based on measurements of the upstream/downstream pressure and the discharge.

%Note: If a larger value of $x_0$ is chosen, then one would be in effect averaging the porous-medium resistivity with the vanishing resistivity of the empty buffer region.

%-----------------------------------------------------------------------------------------------%
\subsection{Relationship to total drag}

The resistivity is directly related to the drag exerted by the bodies. Indeed, another way to express the Stokes equations is $\grad \cdot \bvec{\sigma} = 0$ where $\bvec{\sigma} = -p \bvec{I} + \mu \left( \grad \bu + \grad \bu^T \right)$ is the stress tensor. The divergence theorem gives
\begin{equation}
\label{divthm}
0 = \int_{\Omega} \grad \cdot \bvec{\sigma} \, dV 
= \int_{\partial \Omega} \bvec{\sigma} \cdot \nhat \, dS
\end{equation}
for any domain $\Omega$.  Consider $\Omega$ to be inside $[-x_0, x_0] \times [-1, 1]$, where we will choose $x_0 > 1$ so that $\Omega$ includes the entire porous region.
The boundary integral on the right-hand-side of \eqref{divthm} has contributions from the boundaries of the erodible bodies $\gamma$ and from the outer boundary $\Gamma$, since $\partial \Omega = \gamma \cup \Gamma$. In particular, the total drag exerted by the erodible bodies is given by
\begin{equation}
\FD = (D_1, D_2) = -\int_{\gamma} \bvec{\sigma} \cdot \nhat \, dS 
\end{equation}
which includes both pressure and viscous drag. The minus sign is due to the orientation of the normal vector. We will mostly be interested in the horizontal component $D_1$.

Meanwhile, the horizontal component of the other term is
\begin{align}
\ex \cdot \int_{\Gamma} \bvec{\sigma} \cdot \nhat \, dS = 
& \int_{-1}^{1} p(-x_0, y) - p(x_0, y) - 2\mu (u_x(-x_0,y) - u_x(x_0,y)) dy \, + \\
& \int_{-x_0}^{x_0} \mu (u_y(x,1) - u_y(x,-1)) dx
\end{align}
If we take $x_0$ sufficiently far away, then the viscous terms involving $u_x$ will become small as the velocity field tends to a uniform flow. Meanwhile, the contribution from the pressure is well approximated by the upstream and downstream average pressures $\pup$ and $\pdn$ as introduced in the previous section. This gives 
\begin{equation}
D_1 \approx 2 (\pup - \pdn)
\end{equation}

Therefore we have the relationship
\begin{equation}
r = \frac{D_1}{4 \qavg}
\end{equation}
